---
title: "PM566 Final Project"
author: "Yiping Li"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

<br>

This is my PM566 Final Project website.

The pdf version of the Final Project can be found at: https://github.com/marylee513/pm566_final_project/blob/main/PM566%20Final%20Project.pdf

Introduction: 
In the United States, more than half of the total dietary energy consumption has been composed by ultra-processed foods (UPFs). Most UPFs are calorie-dense and high in sugar, salt, and unhealthy fats, while low in protein, vitamins, and minerals. Studies have demonstrated that the increased consumption of UPFs could result in a poor nutritional quality of the overall diet and high chances of developing chronic diseases. Therefore, the aim of the study focuses on exploring the association between UPF consumption and type 2 diabetes (T2DM) and obesity among youth. 

<br>

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(readxl)
library(plyr)
library(dplyr)
library(skimr)
library(ggplot2)
library(plotly)
library(knitr)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
  class.source = "code-r")
```

```{css, echo = FALSE}
.code-r { /* Code block */
  font-size: 15px;
}

.code-r-small { /* Code block */
  font-size: 10px;
}
```

<br>

## First Step

Firstly, I sourced codes from "Midterm.R".

```{r load-data}
source("Midterm.R")
```

<br>

## Second Step

Secondly, I copied and pasted codes needed from source.

```{r fig1 to fig3}
fig1 <- 
ggplot(vars, aes(x = UPFpercentage_mean, y = BMI_cont)) + 
  geom_point() +
  labs(title = "Fig1: Scatterplot of UPF vs. BMI", x = "UPF%", y = "BMI")

fig2 <- 
ggplot(vars, aes(x = UPFpercentage_mean, y = Gluc_Fasting)) + 
  geom_point() +
  labs(title = "Fig2: Scatterplot of UPF vs. Fasting Glucose", x = "UPF%", y = "Fasting Glucose")

fig3 <- 
ggplot(vars, aes(x = UPFpercentage_mean, y = Gluc_120min)) + 
  geom_point() +
  labs(title = "Fig3: Scatterplot of UPF vs. Blood Glucose after 120 minutes", x = "UPF%", y = "Blood Glucose after 120 minutes")
```

<br>

## Third Step {.tabset}

I created tabs to display each figure using function ggplotly.

Figure 1 to 3 shows UPF% vs BMI, fasting glucose, and blood glucose after 120 minutes correspondingly. None of the plots show a trend of UPF% on those variables interested in. 

### Fig1
```{r fig1}
ggplotly(fig1)
```

### Fig2
```{r fig2}
ggplotly(fig2)
```

### Fig3
```{r fig3}
ggplotly(fig3)
```

## {-}
